---
title: "Case Study 2: How Can a Wellness Technology Compabny Play it Smart?"
author: "Grant G"
date: '2022-07-11'
output: html_document
---

# **Ask Phase**


### The business task is to find trends in the provided fitness tracking application to help inform the Bellabeat marketing team on how to target their consumers. This includes data related to the number of total steps, intensity of the exercise, and sleep time for each of the 30 participants in the study. The focus for the study is on a particular product produced by the company Bellabeat: Bellabeat app. By understanding the lifestyle patterns of women, Bellabeat can help people improve their lives through healthy changes in their everyday routines. 


# **Stakeholders**



## Key Stakeholders Include:

### CCO Urška Sršen
### Cofounder and executive team member Sando Mur 
### The analytics team of Bellabeat 







# **Prepare Phase**

### Data for this project is located on the website Kaggle through the company Mobius. A variety of files are found listing fitness tracking data [here](https://www.kaggle.com/datasets/arashnic/fitbit).

### The data found on Kaggle is considered to be open source. Therefore, concerns surrounding the privacy of data is limited and all permissions have been granted to the public for manipulation of the data for personal purposes. 

### The data given has been recorded from March 12/2016-May 12/2016. This data differs in format based on the specific brand of fitness tracker used by the participant in the study. Variation will also occur based on personal use and consistency of ensuring the product is worn at all times of the day while awake. 

### Potentional concerns of the dataset include but are not limited too: The sample size of the study (30 participants) being too low to accurately represent the population as a whole, errors with tracking device accurately measuring caloric burn and number of steps taken (caloric burn is based on a calculation made by the Bellabeat application for a particular age, weight and sex). Step count uses an excelerometer found inside the device which could count false readings while performing movements that don't involve running or walking)


# **Process Phase** 


## Installation of the appropriate packages 

```{r}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("skimr")
install.packages("janitor")
install.packages("here")
install.packages("lubridate")
install.packages("ggpubr")
install.packages("ggrepel")
```


## Loading of the installed packages by opening the libraries 



```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(skimr)
library(janitor)
library(here)
library(lubridate)
library(ggpubr)
library(ggrepel)
```


## The appropriate files were imported into the RStudio Cloud platform from Kaggle



```{r}
daily_activity <- read_csv(file= "/cloud/project/Bellabeat.csv")
daily_sleep <- read_csv(file= "/cloud/project/sleepDay_Merged.csv")
hourly_steps <- read_csv("/cloud/project/hourlySteps_merged.csv")
```




## Previewing the files

```{r}

head(daily_activity)
str(daily_activity)


head(daily_sleep)
str(daily_sleep)


head(hourly_steps)
str(hourly_steps)


```



# **Cleaning of the Data**


## The date times were partially formatted in excel to be read correctly by RStudio Cloud
## Looking for the number of unique Id's in each of the data sets

```{r}
n_unique(daily_activity$Id)
n_unique(daily_sleep$Id)
n_unique(hourly_steps$Id)
```


## Searching each of the data sets for duplicates 

```{r}
sum(duplicated(daily_activity))
sum(duplicated(daily_sleep))
sum(duplicated(hourly_steps))
```


## Remove the duplicated files from each dataset

```{r}
daily_activity <- daily_activity %>%
  distinct() %>%
  drop_na()

daily_sleep <- daily_sleep %>%
  distinct() %>%
  drop_na()

hourly_steps <- hourly_steps %>%
  distinct() %>%
  drop_na()

```


##  Ensure that all duplicates have been removed from the dataset 

```{r}
sum(duplicated(daily_activity))
sum(duplicated(daily_sleep))
sum(duplicated(hourly_steps))
```


##  Ensure all column names are formatted to be the same ( lower case for all columns and only characters, numbers and underscores)

```{r}
clean_names(daily_activity)
daily_activity<- rename_with(daily_activity, tolower)
clean_names(daily_sleep)
daily_sleep <- rename_with(daily_sleep, tolower)
clean_names(hourly_steps)
hourly_steps <- rename_with(hourly_steps, tolower)

```


## Set Consistent Date and Time Columns ( )

```{r}
daily_activity <- daily_activity %>%
  rename(date = activity_date) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

daily_sleep <- daily_sleep %>%
  rename(date = sleepday) %>%
  mutate(date = as_date(date,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))



hourly_steps<- hourly_steps %>% 
  rename(date_time = activityhour) %>% 
  mutate(date_time = as.POSIXct(date_time,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))
```



## Merging Datasets ( Merging daily_activity with daily_sleep where id and date are the same in both data sets)

```{r}
daily_activity_sleep <- merge(daily_activity, daily_sleep, by=c ("id", "date"))
glimpse(daily_activity_sleep)

```


# **Analyze Phase and Share Phase** 


### In order to understand the overall level of activity exhibited by each person, we subcategorize each participant based on the average step count per day. These categorizations are based on reccomendations from healthline [click here](https://www.healthline.com/health/how-many-steps-a-day#Why-10,000-steps?)



### Arrange the chart in descending order to show which people have the most steps compared to others in the group and to search for outliers in the data relative to the average number of total steps ( Person "8877689391 " Has a tendency to walk more steps by a large margin in comparison to other people. This is also seen in the Very_Active_Distance where they rank 5 times in the top 10 of the overall data)

```{r}
Total_Steps_DESC <-arrange(daily_activity,-total_steps)
Very_Active_Steps_DESC <- arrange(daily_activity,-very_active_distance)



```


## Calculate the average total steps, calories burned, average sleep by each person 

```{r}
daily_average <- daily_activity_sleep %>%
  group_by(id) %>%
  summarise (mean_daily_steps = mean(total_steps), mean_daily_calories = mean(calories), mean_daily_sleep = mean(totalminutesasleep))


```

## Calculating the mean distance by level of intensity for each day of the week  

```{r}
Average_Steps_Day_Intensity <-
  daily_activity %>%
  group_by(day_of_the_week)%>% 
  summarize( Average_Total_Distance = mean(total_steps),
             very_mean = mean(very_active_distance),
             moderate_mean = mean(moderately_active_distance),
             light_mean = mean(light_active_distance))
```


## Classify our users based on daily average steps. As per Healthline website, fewer than 5000 steps is considered low, while the ideal number of steps per day should approach 10000.

### Average steps less than 5000 steps is considered sedentary
### Average steps between 5000 and 7499 steps is considered lightly active
### Average steps between 7500 steps and 10000 steps is considered fairly active
### Average steps between 10000 steps and 12499 steps is considered very active
### Average steps greater than 12500 is considered extremely active 

```{r}
user_type <- daily_average %>%
  mutate(user_type = case_when(
    mean_daily_steps < 5000 ~ "sedentary",
    mean_daily_steps >= 5000 & mean_daily_steps < 7499 ~ "lightly active", 
    mean_daily_steps >= 7500 & mean_daily_steps < 9999 ~ "fairly active", 
    mean_daily_steps >= 10000 & mean_daily_steps < 12499 ~ "very active",
    mean_daily_steps >= 12500 ~ "extremely active"
  ))

```


## Finding the percentage of each level of activity based on the whole

```{r}
user_type_percent <- user_type %>%
  group_by(user_type) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(user_type) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))

user_type_percent$user_type <- factor(user_type_percent$user_type , levels = c("extremely active", "very active", "fairly active", "lightly active", "sedentary"))

```


## Creation of Pie chart based on levels of activity for each individual relative to the whole 

### The graph below demonstrates that most people are able to achieve an appropriate amount of exercise each day 

```{r}
user_type_percent %>%
  ggplot(aes(x="",y=total_percent, fill=user_type)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=6)+
  theme_minimal()+
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(family = "TT Times New Roman", hjust = 0.8, vjust = 0, size=16, face = "bold.italic")) +
  scale_fill_manual(values = c("#000EFF","#D3FF00", "#00DBFF", "#FF0000", "#15E93D")) +
  geom_text(aes(label = labels),position = position_stack(vjust = 0.5))+
  labs(title="Users by Activity Level")
```




## Day of the week relative to intensity of exercise

### This data will compare the intensity of exercise performed based on each day of the week. This will indicate how much of the overall distance each person completes is based primarily on one type of exercise intensity over another form. The graph demonstrates that intensity of exercise is concentrated on Saturday and Tuesday for moderate and light exercise.

```{r}
ggarrange(
  ggplot(Average_Steps_Day_Intensity) +
    geom_col(aes(day_of_the_week , very_mean), fill = "#FF0000") +
    labs(title = "Average Distance Very", x= "", y = "Average Distance") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1)),
  ggplot(Average_Steps_Day_Intensity, aes(day_of_the_week , moderate_mean)) +
    geom_col(fill = "#000EFF") +
    labs(title = "Average Distance Moderate", x= "", y = "Average Distance") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1)),
  ggplot(Average_Steps_Day_Intensity, aes(day_of_the_week , light_mean)) +
    geom_col(fill = "#DFF212") +
    labs(title = "Average Distance Light", x= "", y = "Average Distance") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1))
)
```


## Hourly Steps Throughout the Day


### The graph below demonstrates that the majority of steps are completed between the hours of 07:00-18:00

```{r}
hourly_steps <- hourly_steps %>%
  separate(date_time, into = c("date", "time"), sep= " ") %>%
  mutate(date = ymd(date)) 

hourly_steps %>%
  group_by(time) %>%
  summarize(average_steps = mean(steptotal)) %>%
  ggplot() +
  geom_col(mapping = aes(x=time, y = average_steps, fill = average_steps)) + 
  labs(title = "Hourly steps throughout the day", x="", y="") + 
  scale_fill_gradient(low = "purple", high = "red")+
  theme(axis.text.x = element_text(angle = 45))

```






##  **Correlations**

### These graphs will plot the number of step vs total minutes slept and total sleep vs calories burned in one day 

### The graphs below demonstrate that the number of minutes slept in not correlated with the total number of steps taken while the number of calories burned in positively corelated with the number of steps taken. 

```{r}
ggarrange(
  ggplot(daily_activity_sleep, aes(x=total_steps, y=totalminutesasleep))+
    geom_jitter() +
    geom_smooth(color = "red") + 
    labs(title = "Daily steps vs Minutes asleep", x = "Daily steps", y= "Minutes asleep") +
    theme(panel.background = element_blank(),
          plot.title = element_text( size=14)), 
  ggplot(daily_activity_sleep, aes(x=total_steps, y=calories))+
    geom_jitter() +
    geom_smooth(color = "red") + 
    labs(title = "Daily steps vs Calories", x = "Daily steps", y= "Calories") +
    theme(panel.background = element_blank(),
          plot.title = element_text( size=14))
)
```




## Number of days each person uses their device to track health data

### Participants are seperated into three different groups based on the use through the 30 day period

```{r}
daily_use <- daily_activity_sleep %>%
  group_by(id) %>%
  summarize(days_used=sum(n())) %>%
  mutate(usage = case_when(
    days_used >= 1 & days_used <= 6 ~ "very low use",
    days_used >= 7 & days_used <= 12 ~ "low use",
    days_used >= 13 & days_used <= 18 ~ "moderate use", 
    days_used >= 19 & days_used <= 24 ~ "high use", 
    days_used >= 25 & days_used <= 31 ~ "very high use"
  ))
```




## Group by overall usage of the the device and find the percentage of each part of the whole 
```{r}
daily_use_percent <- daily_use %>%
  group_by(usage) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(usage) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))
```


## Use of the factor() function to categorize and store the data as vector integer values 

```{r}
daily_use_percent$usage <- factor(daily_use_percent$usage, levels = c("very high use" , "high use", "moderate use", "low use" , "very low use"))
```



## Plot of the daily use of smart device based on number of days used in a month 

### The pie chart demonstrates that most people use their device consistently throughout the month 

```{r}
daily_use_percent %>%
  ggplot(aes(x="",y=total_percent, fill=usage)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))+
  scale_fill_manual(values = c( "#1A3552", "#28517D","#366AA2","#4585CB","#55A7FF" ),
                    labels = c("Very High use - 25 to 31 days",
                               "High use - 19 to 24 days",
                               "Moderate use - 13 to 18 days",
                               "Low use - 7 to 12 days",
                               " Very Low use - 1 to 6 days"))+
  labs(title="Daily use of smart device")
```


## Merging of the daily_activity and daily_use data sets based on the category id 

### We will now be looking at how consistently each person is using their smart device throughout the day 


```{r}
daily_use_merged <- merge(daily_activity, daily_use, by=c ("id"))
```



## Creating a chart that breaks down the amount of time in minutes each person wears their smart device 


```{r}

minutes_worn <- daily_use_merged %>% 
  mutate(total_minutes_worn = very_active_minutes+fairly_active_minutes+lightly_active_minutes+sedentaryminutes)%>%
  mutate (percent_minutes_worn = (total_minutes_worn/1440)*100) %>%
  mutate (worn = case_when(
    percent_minutes_worn == 100 ~ "All day",
    percent_minutes_worn < 100 & percent_minutes_worn >= 50~ "More than half day", 
    percent_minutes_worn < 50 & percent_minutes_worn > 0 ~ "Less than half day"
  ))
```



## Finding the percentage of the whole for overall wear in each day

```{r}
minutes_worn_percent<- minutes_worn%>%
  group_by(worn) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(worn) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))
```




## Finding the percentage of the whole grouping by high use 

```{r}
minutes_worn_highuse <- minutes_worn%>%
  filter (usage == "high use")%>%
  group_by(worn) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(worn) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))
```


## Finding the percentage of the whole grouping by moderate use

```{r}

minutes_worn_moduse <- minutes_worn%>%
  filter(usage == "moderate use") %>%
  group_by(worn) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(worn) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))


```



## Finding the percentage of the whole grouping by low use


```{r}
minutes_worn_lowuse <- minutes_worn%>%
  filter (usage == "low use") %>%
  group_by(worn) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(worn) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))
```




## Use of the factor() function to categorize and store the data as vector integer values 

```{r}
minutes_worn_highuse$worn <- factor(minutes_worn_highuse$worn, levels = c("All day", "More than half day", "Less than half day"))
minutes_worn_percent$worn <- factor(minutes_worn_percent$worn, levels = c("All day", "More than half day", "Less than half day"))
minutes_worn_moduse$worn <- factor(minutes_worn_moduse$worn, levels = c("All day", "More than half day", "Less than half day"))
minutes_worn_lowuse$worn <- factor(minutes_worn_lowuse$worn, levels = c("All day", "More than half day", "Less than half day"))
```


# Plotting of the minutes data 

```{r}
ggarrange(
  ggplot(minutes_worn_percent, aes(x="",y=total_percent, fill=worn)) +
    geom_bar(stat = "identity", width = 1)+
    coord_polar("y", start=0)+
    theme_minimal()+
    theme(axis.title.x= element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(), 
          panel.grid = element_blank(), 
          axis.ticks = element_blank(),
          axis.text.x = element_blank(),
          plot.title = element_text(hjust = 0.5, size=14, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5)) +
    scale_fill_manual(values = c("#004d99", "#3399ff", "#cce6ff"))+
    geom_text(aes(label = labels),
              position = position_stack(vjust = 0.5), size = 3.5)+
    labs(title="Time worn per day", subtitle = "Total Users"),
  ggarrange(
    ggplot(minutes_worn_highuse, aes(x="",y=total_percent, fill=worn)) +
      geom_bar(stat = "identity", width = 1)+
      coord_polar("y", start=0)+
      theme_minimal()+
      theme(axis.title.x= element_blank(),
            axis.title.y = element_blank(),
            panel.border = element_blank(), 
            panel.grid = element_blank(), 
            axis.ticks = element_blank(),
            axis.text.x = element_blank(),
            plot.title = element_text(hjust = 0.5, size=14, face = "bold"),
            plot.subtitle = element_text(hjust = 0.5), 
            legend.position = "none")+
      scale_fill_manual(values = c("#004d99", "#3399ff", "#cce6ff"))+
      geom_text_repel(aes(label = labels),
                      position = position_stack(vjust = 0.5), size = 3)+
      labs(title="", subtitle = "High use - Users"), 
    ggplot(minutes_worn_moduse, aes(x="",y=total_percent, fill=worn)) +
      geom_bar(stat = "identity", width = 1)+
      coord_polar("y", start=0)+
      theme_minimal()+
      theme(axis.title.x= element_blank(),
            axis.title.y = element_blank(),
            panel.border = element_blank(), 
            panel.grid = element_blank(), 
            axis.ticks = element_blank(),
            axis.text.x = element_blank(),
            plot.title = element_text(hjust = 0.5, size=14, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none") +
      scale_fill_manual(values = c("#004d99", "#3399ff", "#cce6ff"))+
      geom_text(aes(label = labels),
                position = position_stack(vjust = 0.5), size = 3)+
      labs(title="", subtitle = "Moderate use - Users"), 
    ggplot(minutes_worn_lowuse, aes(x="",y=total_percent, fill=worn)) +
      geom_bar(stat = "identity", width = 1)+
      coord_polar("y", start=0)+
      theme_minimal()+
      theme(axis.title.x= element_blank(),
            axis.title.y = element_blank(),
            panel.border = element_blank(), 
            panel.grid = element_blank(), 
            axis.ticks = element_blank(),
            axis.text.x = element_blank(),
            plot.title = element_text(hjust = 0.5, size=14, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none") +
      scale_fill_manual(values = c("#004d99", "#3399ff", "#cce6ff"))+
      geom_text(aes(label = labels),
                position = position_stack(vjust = 0.5), size = 3)+
      labs(title="", subtitle = "Low use - Users"), 
    ncol = 3), 
  nrow = 2)

```

### The data demonstrates these conclusions:

### Most people will consistently use their smart device at least half of the day or more 
### For High users, the device is primarily used more than half of the day where a minority group will wear the device all day
### For Moderate users, there is an increase in the number of participants that use the device all day and decrease from half day or more in comparison to High users
### For Low users, The device is almost exclusively worn all day 


# **Conclusion**

## Based on the provided correlations and data visualizations, their are a variety of potential connections that can be made to further the health of women. The limitations of small sample size and improper use of the smart device makes providing concrete conclusions difficult. None the less, here are a few possible takeaways from the data:

### 1. Many of the participants were inconsistently using their smart device to track varying health related metrics. Notifications should be used through changes in personal lifestyle habits such as writing down a routine in the morning or setting up automated notifications in other devices such as a smart phone. This will allow the participants to know whether they have reached the reccommended step count and sleep time each day

### 2. Many of the participants are attempting to achieve their step count through short bursts of vigorous exercise. This may be useful if their occupation requires them to remain sedentary for long periods of time. However, finding time to exercise can be difficult for many people and therefore spreading out exercise throughout the day through small walking breaks could alleviate this pressure

### 3. Participants should also consider changes to their bedtime routine to increase the amount of overall sleep they acheive. Many particpants did not acheive the required 8 hours of sleep per day as recommended by the CDC. Also placing smart devices away from the bed and decrease the exposure to electronic devices one hour before bed could help increase the quality and length of sleep of participants. 
